---
output: html_document
editor_options: 
  chunk_output_type: inline
---

Red Wine Investigation by Russell Harris
========================================================

#Dataset Description
> The red wine dataset is used for exploring chemical properties of red wine to 
see what factors correlates to wine quality ratings.
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
Modeling wine preferences by data mining from physicochemical properties. In 
Decision Support Systems, Elsevier, 47(4):547-553, 2009.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

# reset Environment
#rm(list=ls())

# install these if necessary
#install.packages('dplyr')
#install.packages('GGally')
#install.packages('ggplot2')
#install.packages('gridExtra')
#install.packages('lattice')
#install.packages('MASS')
#install.packages('memisc')
#install.packages('reshape2')

# load the ggplot graphics package and the others
library(dplyr)
library(GGally)
library(ggplot2)
library(grid)
library(gridExtra)
library(lattice)
library(MASS)
library(memisc)
library(reshape2)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
#df <- read.csv(file.choose())
setwd('H:/Udacity/DAND/Term 2/1-Exploratory Data Analysis/Project')
df <- read.csv('wineQualityReds.csv')
df$quality.factor <- as.factor(df$quality)

#Data Structure
str(df)
```
>After analyzing the data, a new variable called 'quality.factor' was created \
to support further analysis of the data.
```{r echo=FALSE}
#View Data in seperate window
#View(df)

#Data Summary
summary(df)

#Red Wine by rating (3-8)
#table(df$quality)
```
> * The data summary shows 1599 samples in the dataset.
* Actual Wine quality ratings range from 3-8, even though 0-10 is possible.

# Univariate Plots Section
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality), data=df)+
  geom_bar(color =I('black'),fill = I('forestgreen'))+
  ggtitle('Wine Quality Distribution (barchart)')
```

> The distrubtion of quality is a normal distribution. From the data summary, \
the count in each quality rating is:\
3: 10\
4: 53\
5: 681\
6: 638\
7: 199\
8: 18

```{r echo=FALSE, message=FALSE, warning=FALSE}
#http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
p1<-ggplot(aes(x=fixed.acidity), data=df)+ 
  geom_histogram(color =I('black'),fill = I('darkblue'))+
  #ggtitle('Fixed Acidity')+
  xlab('fixed acidity')

p2<-ggplot(aes(x=volatile.acidity), data=df) +
  geom_histogram(color =I('black'),fill = I('darkcyan')) +
  #ggtitle('Volatile Acidity')+
  xlab('volatile acidity')

p3<-ggplot(aes(x=citric.acid), data=df) +
  geom_histogram(color =I('black'),fill = I('darkgoldenrod')) +
  #ggtitle('Citric Acid')+
  xlab('citric acid')

p4<-ggplot(aes(x=residual.sugar), data=df)+
  geom_histogram(color =I('black'),fill = I('firebrick'))+
  #ggtitle('Residual Sugar')+
  xlab('residual sugar')

p5<-ggplot(aes(x=chlorides), data=df)+
  geom_histogram(color =I('black'),fill = I('darkkhaki'))
  #ggtitle('Chlorides')

p6<-ggplot(aes(x=free.sulfur.dioxide), data=df)+
  geom_histogram(color =I('black'),fill = I('darkmagenta'))+
  #ggtitle('Free Sulfur Dioxide')+
  xlab('free sulfur dioxide')

p7<-ggplot(aes(x=total.sulfur.dioxide), data=df)+
  geom_histogram(color =I('black'),fill = I('darkolivegreen'))+
  #ggtitle('Total Sulfur Dioxide')+
  xlab('total sulfur dioxide')

p8<-ggplot(aes(x=density), data=df)+
  geom_histogram(color =I('black'),fill = I('goldenrod'))
  #ggtitle('Density')

p9<-ggplot(aes(x=pH), data=df)+
  geom_histogram(color =I('black'),fill = I('darkorchid'))
  #ggtitle('pH')

p10<-ggplot(aes(x=sulphates), data=df)+
  geom_histogram(color =I('black'),fill = I('darkred'))
  #ggtitle('Sulphates')

p11<-ggplot(aes(x=alcohol), data=df)+
  geom_histogram(color =I('black'),fill = I('darksalmon'))
  #ggtitle('Alcohol')

#p12<-ggplot(aes(x=quality), data=df)+
#  geom_histogram(color =I('black'),fill = I('forestgreen'))
#  #ggtitle('Quality')

grid.arrange(p8,p9,ncol=2,
  top = textGrob("Wine Attributes: Normally Distributed",gp=gpar(fontsize=14)))
```

> Normally Distributed: Density and pH.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(p1,p2,p3,p6,p7,p10,p11,ncol=4,
  top = textGrob("Wine Attributes: Right-Skewed",gp=gpar(fontsize=14)))
```

> Right-Skewed: Fixed acidity, volatile acidity, free/total sulfur dioxide, \
sulphates, and alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(p4,p5,ncol=2,
  top = textGrob("Wine Attributes: With large outliers",gp=gpar(fontsize=14)))
```

> Attributes with Extreme Outliers are: Residual sugar, chlorides \
Boxplots are the best graphing choice to view which variables have outliers \
that could impact the data anaysis.

```{r echo=FALSE, message=FALSE, warning=FALSE}
s1<-ggplot(aes(y=fixed.acidity), data=df)+ 
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')+
  ylab('fixed acidity')

s2<-ggplot(aes(y=volatile.acidity), data=df) +
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')+
  ylab('volatile acidity')

s3<-ggplot(aes(y=citric.acid), data=df) +
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')+
  ylab('citric acid')

s4<-ggplot(aes(y=residual.sugar), data=df)+
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')+
  ylab('residual sugar')

s5<-ggplot(aes(y=chlorides), data=df)+
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')

s6<-ggplot(aes(y=free.sulfur.dioxide), data=df)+
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')+
  ylab('free sulfur dioxide')

s7<-ggplot(aes(y=total.sulfur.dioxide), data=df)+
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')+
  ylab('total sulfur dioxide')

s8<-ggplot(aes(y=density), data=df)+
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')

s9<-ggplot(aes(y=pH), data=df)+
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')

s10<-ggplot(aes(y=sulphates), data=df)+
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')

s11<-ggplot(aes(y=alcohol), data=df)+
  geom_boxplot(colour = "black", fill = I('lightskyblue'), 
               outlier.color = 'red')

grid.arrange(s1,s2,s3,s4,s5,s6,s7,s7,s9,s10,s11,ncol=4, 
  top = textGrob("Wine Attributes: Boxplot",gp=gpar(fontsize=14)))
```

> From the boxplots, we can see that resudual sugar and \
chlorides definitely have large outlier data. We'll graph just those next.

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(s4,s5,ncol=2, 
  top = textGrob("Wine Attributes with HUGE Outliers: Boxplot",gp=gpar(fontsize=14)))
```

> The Boxplots show that both residual siguar and chlorides have large \
outliers. It could be useful to remove some of the extreme data points to get \
a better analysis.

# Univariate Plots Observations
>* Normally Distributed: Density and pH.
* Right-Skewed: Fixed acidity, volatile acidity, free/total sulfur dioxide, \
sulphates, citric acid, and alcohol
* Has Extreme Outliers: Residual sugar, chlorides

### What is the structure of your dataset?
> The red wine dataset has 1599 observations of 13 numeric variables with an \
integer rating between 0-10 possible (actual is 3-8).

### What is/are the main feature(s) of interest in your dataset?
> The variables that correlate the most to the wine rating is of most interest.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
> If the normally distributed variables or right-skewed variables have a \
bigger impact to quality.

### Did you create any new variables from existing variables in the dataset?
> Yes, I created a new factor variable from quality called quality.factor.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
> No, all data was left in its original form.


# Bivariate Plots Section

# Variable to Quality Correlation
> Before creating our bivariate plots, let's check The correlation coefficients\
 of quality to each wine attribute.
 
```{r echo=FALSE, Bivariate_Analysis}
correlations <- c(
  cor(df$fixed.acidity, df$quality),
  cor(df$volatile.acidity, df$quality),
  cor(df$citric.acid, df$quality),
  cor(df$residual.sugar, df$quality),
  cor(df$chlorides, df$quality),
  cor(df$free.sulfur.dioxide, df$quality),
  cor(df$total.sulfur.dioxide, df$quality),
  cor(df$density, df$quality),
  cor(df$pH, df$quality),
  cor(df$sulphates, df$quality),
  cor(df$alcohol, df$quality)
)

names(correlations) <- c('fixed acidity',
                         'volatile acidity',
                         'citric acid',
                         'residual sugar',
                         'chlorides',
                         'free sulfur dioxide',
                         'total sulfur dioxide',
                         'density',
                         'pH',
                         'sulphates',
                         'alcohol'
                         )

correlations[order(abs(correlations), decreasing = TRUE)]
```
> Quality is most correlated to the following four attributes (in order):\
 1. alcohol\
 2. volatile acidity\
 3. sulphates\
 4. citric acid

```{r echo=FALSE, Bivariate_Plots}
#Source: http://www.cookbook-r.com/Graphs/Scatterplots_(ggplot2)/
ggplot(data=df, aes(x=quality.factor, y=alcohol))+
  geom_jitter(alpha=0.1)+
  geom_boxplot(alpha = 0.1, color = 'blue') +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red') +
  labs(x = 'Quality', y = 'Alcohol',
  title = 'Boxplot of Alcohol Across Quality')
```

> Quality goes up when alcohol goes up.

```{r echo=FALSE}
ggplot(data=df, aes(x=quality.factor, y=volatile.acidity))+
  geom_jitter(alpha=0.1)+
  geom_boxplot(alpha = 0.1, color = 'blue') +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red') +
  labs(x = 'Quality', y = 'volatile.acidity',
  title = 'Boxplot of Volatile Acidity Across Quality')
```

> Quality goes up when volatile acidity goes down.

```{r echo=FALSE}
#Source: http://www.cookbook-r.com/Graphs/Scatterplots_(ggplot2)/
ggplot(data=df, aes(x=quality.factor, y=sulphates))+
  geom_jitter(alpha=0.1)+
  geom_boxplot(alpha = 0.1, color = 'blue') +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red') +
  labs(x = 'Quality', y = 'sulphates',
  title = 'Boxplot of Sulphates Across Quality')
```

> Quality goes up when sulphates goes up.

```{r echo=FALSE}
#Source: http://www.cookbook-r.com/Graphs/Scatterplots_(ggplot2)/
ggplot(data=df, aes(x=quality.factor, y=citric.acid))+
  geom_jitter(alpha=0.1)+
  geom_boxplot(alpha = 0.1, color = 'blue') +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red') +
  labs(x = 'Quality', y = 'citric.acid',
  title = 'Boxplot of Citric Acid Across Quality')
```

> Quality goes up as citric acid goes up.\
\
As expected from the correlation coefficient calculations in the boxplots, \
we see that alcohol, sulphates, citric acid are positively correlated to \
quality. While volatile acidity is negatively correldated to quality.

# Alcohol to other variables correlation
> Since alcohol is the biggest factor for predicting quality, let's check \
how other variables are correlated with alcohol.

```{r echo=FALSE}
correlations <- c(
  cor(df$fixed.acidity, df$alcohol),
  cor(df$volatile.acidity, df$alcohol),
  cor(df$citric.acid, df$alcohol),
  cor(df$residual.sugar, df$alcohol),
  cor(df$chlorides, df$alcohol),
  cor(df$free.sulfur.dioxide, df$alcohol),
  cor(df$total.sulfur.dioxide, df$alcohol),
  cor(df$density, df$alcohol),
  cor(df$pH, df$alcohol),
  cor(df$sulphates, df$alcohol)
)

names(correlations) <- c('fixed acidity',
                         'volatile acidity',
                         'citric acid',
                         'residual sugar',
                         'chlorides',
                         'free sulfur dioxide',
                         'total sulfur dioxide',
                         'density',
                         'pH',
                         'sulphates'
                         )

correlations[order(abs(correlations), decreasing = TRUE)]
```
> Top correlations to alcohol:\
1) density\
2) chlorides\
3) total sulfur dioxide\
4) pH\
5) volatile acidity

```{r echo=FALSE}
ggplot(data=df, aes(x=density, y=alcohol))+
  geom_jitter(alpha=0.2)+
  geom_smooth(method=lm, se=TRUE)+
  labs(x = 'Density', y = 'Alcohol',
  title = 'Scatterplot of Density and Alcohol')
```

> Alcohol increases as density decreases

```{r echo=FALSE}
t1<-ggplot(data=df, aes(x=chlorides, y=alcohol))+
  geom_jitter(alpha=0.2)+
  geom_smooth(method=lm, se=TRUE)+
  labs(x = 'Chlorides', y = 'Alcohol')

t2<-ggplot(data=df, aes(x=total.sulfur.dioxide, y=alcohol))+
  geom_jitter(alpha=0.2)+
  geom_smooth(method=lm, se=TRUE)+
  labs(x = 'Total Sulfur Dioxide', y = 'Alcohol')

t3<-ggplot(data=df, aes(x=pH, y=alcohol))+
  geom_jitter(alpha=0.2)+
  geom_smooth(method=lm, se=TRUE)+
  labs(x = 'pH', y = 'Alcohol')

t4<-ggplot(data=df, aes(x=volatile.acidity, y=alcohol))+
  geom_jitter(alpha=0.2)+
  geom_smooth(method=lm, se=TRUE)+
  labs(x = 'Volatile Acidity', y = 'Alcohol')
grid.arrange(t1,t2,t3,t4,ncol=2, 
  top = textGrob("Other Alcohol Attributes Scatterplots",gp=gpar(fontsize=14)))
```

> * Chlorides looks to cluster and not have a good visual correlation to \
alcohol.\
 * Alcohol decreases as total sulfur dioxide and volatile acidity decrease.\
 * Alcohol increases as pH increases.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
> From the bivariate plots and correlations, we can see the quality is most \
correlated to the following four attributes (in order):\
 * alcohol\
 * volatile acidity\
 * sulphates\
 * citric acid

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
> While total sulfur dioxide had a small correlation, free sulfur dioxide had \
almost no correlation to quality rating.

### What was the strongest relationship you found?
> Ironically, alcohol has the biggest correlation to quality rating.

# Multivariate Plots Section
> For multivariate analysis, I once again took the top four correlated \
attributes (alcohol, volatile acidity, sulphates, citric acid), and \
analyzed them against quality.

```{r echo=FALSE, Multivariate_Plots}
ggplot(data=df, aes(x=volatile.acidity, y=alcohol,
      color = quality.factor)) + 
      geom_jitter() +
      coord_cartesian(xlim = c(0.1, 1.2), ylim = c(9, 14)) +
      scale_color_brewer(type = 'seq') +
      theme_dark() +
      labs(x='volatile acidity', y='alcohol', colour='quality',
      title='Alcohol VS Volatile Acidity with Colored Quality')
```

> Quality increases when alcohol increases and volatile acidity \
decreases.

```{r echo=FALSE}
ggplot(data=df, aes(x=sulphates, y=alcohol,
      color = quality.factor)) + 
      geom_jitter() +
      coord_cartesian(xlim = c(0.3, 1.4), ylim = c(9, 14)) +
      scale_color_brewer(type = 'seq') +
      theme_dark() +
      labs(x='sulphates', y='alcohol', colour='quality',
      title='Alcohol VS Sulphates with Colored Quality')
```

> Quality increases as alcohol and sulphates increase.

```{r echo=FALSE}
ggplot(data=df, aes(x=citric.acid, y=alcohol,
      color = quality.factor)) + 
      geom_jitter() +
      coord_cartesian(xlim = c(0, 0.8), ylim = c(9, 14)) +
      scale_color_brewer(type = 'seq') +
      theme_dark() +
      labs(x='citric acid', y='alcohol', colour='quality',
      title='Alcohol VS Citric Acid with Colored Quality')
```

> Quality increases as alcohol and citric acid increases.

```{r echo=FALSE}
ggplot(data=df, aes(x=density, y=alcohol,
      color = quality.factor)) + 
      geom_jitter() +
      coord_cartesian(xlim = c(0.992, 1.002), ylim = c(9, 14)) +
      scale_color_brewer(type = 'seq') +
      theme_dark() +
      labs(x='density', y='alcohol', colour='quality',
      title='Alcohol VS Density with Colored Quality')
```

> Quality increases as alcohol increases and density decreases.

```{r echo=FALSE}
ggplot(data=df, aes(x=volatile.acidity, y=citric.acid,
      color = quality.factor)) + 
      geom_jitter() +
      coord_cartesian(xlim = c(0.1, 1.1), ylim = c(0, 0.8)) +
      scale_color_brewer(type = 'seq') +
      theme_dark() +
      labs(x='volatile acidity', y='citric acid', colour='quality',
      title='Citric Acid VS Volatile Acidity with Colored Quality')
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
>For the multivariate analysis, I examined the four features that were most \
correlated (alcohol, volatile acidity, sulphates, citric acid) and faceted them\
 by quality. I choose to put citric acid and volatile acid together and it \
 showed that better quality wine was associated with high ctric acide and low \
 volatile acid. Also, the graph confirmed that higher alcohol and higher \
 sulphates resulted in a higher quality wine as predicted in the \
 bivariate analysis section.

### Were there any interesting or surprising interactions between features?
>Not really, the multivariate features visually represented what we found were \
correltated in the bivariate analysis section.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data=df, aes(x=quality.factor, y=alcohol))+
  geom_jitter(alpha=0.1)+
  geom_boxplot(alpha = 0.1, color = 'blue') +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red') +
  labs(x = 'Quality (Rating 0-10)', y = 'Alcohol (% by volume)',
  title = 'Boxplot of Alcohol Across Quality')
```

### Description One
>The boxplots of alcohol (% by volume) show that as it increases, so does \
the rating of the quality of wine. It turns out that alcohol is the biggest
predicter based on correlation coeffiencient.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(data=df, aes(x=density, y=alcohol))+
  geom_jitter(alpha=0.2)+
  geom_smooth(method=lm, se=TRUE)+
  labs(x = 'Density (mg/l)', y = 'Alcohol (% by volume)',
  title = 'Scatterplot of Density and Alcohol')
```

### Description Two
>Alcohol (% by volume) was the biggest predictor of rating, and density (mg/l) \
is the best predictor of alcohol (% by volume). The above graph shows the \
negative correlated relationship between the two variables.


### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(data=df, aes(x=volatile.acidity, y=citric.acid,
      color = quality.factor)) + 
      geom_jitter() +
      coord_cartesian(xlim = c(0.1, 1.1), ylim = c(0, 0.8)) +
      scale_color_brewer(type = 'seq') +
      theme_dark() +
      labs(x='Volatile Acidity (g/dm^3)', y='Citric Acid (g/dm^3)', colour='Quality',
      title='Citric Acid VS Volatile Acidity with Colored Quality')
```

### Description Three
>Volatile acidity (g/dm^3) is negatively correlated to quality, while citric \
acid (g/dm^3) is positively correlated to quality. This multivariate plot \
shows the relationship between these 3 variables perfectly.

------

# Reflection
>The Red Wine Dataset gives a lot of data points to study against overall \
quality ratings. After analyzing all of the variables, it turns out that \
alcohol had the biggest impact on rating. To properly analyze and graph \
everything, I also needed to create a new variable out of quality and set it as a \
factor.\
\
The normal distribution of quality rating in the data made me have more \
confidence in my analysis of the variables and their correlation. In addition, \
the multivariate analysis was one of the most powerful visual tools in seeing \
the different variables and how they impact quality.\
\
My biggest challenge had to do with balancing analysis numerically with \
presenting the data visually. There are numerous ways to customize how you can \
view  data in R from color, zoom, types of plots, and multiple variate analysis\
 approaches. I had to review the course material multiple times to really \
understand the best approach. Ultimately, just going through the different \
analysis levels (univariate, bivariate, multivariate) helped me to progress \
my analysis in a logical order and to better understand my approach. This \
progression is how all data analysis should evolve in the 'real world'.\
\
This was a small sample as there are many more thousands of wine brands \
around the world. I think another point of interest would be seeing if age \
of the wine or price had any correlation to the quality rating. This would \
need to be a blind study as to not bias the testers with any pre-existing \
knowledge.